**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-02 Schloss_Cluster_PeerJ_2015.Rmd
Add commentary on Figure 4

@@

## Results and Discussion
```{r}
z <- read.table(file="data/process/he.mcc_ref.summary", header=T)
```

@@

First, we sought to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset. The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A). Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times. Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception. In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`). This result suggests that the AL algorithm was far more stabile than indicated in the He study. Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.

@@

***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets. These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database because it think it has found the best hit or does not believe that the database contains a hit. This means that the best possible match is not necessarily returned once a highly similar hit is obtained. In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.

* Able to parallelize

* MCC for VAGC/VDGC > AGC/DGC
* MCC for VAGC and VDGC are basically the same
* Stability of VAGC/VDGC goes up

This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.

@@

>>>>

## To Do:
* miseq analysis
* methods section
* references
* figure captions

@@

## Results and Discussion
```{r}
he_stability <- read.table(file="data/process/he.mcc_ref.summary", header=T)
```

@@

First, we sought to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset. The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A). Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times. Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception. In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`). This result suggests that the AL algorithm was far more stabile than indicated in the He study. Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.

@@

***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets. These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database because it think it has found the best hit or does not believe that the database contains a hit. This means that the best possible match is not necessarily returned once a highly similar hit is obtained. In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality. In general the overall trends that we observed with the USEARCH-version of AGC and DGC were also observed with the VSEARCH-verion of the algorithms. When we compared the two implementations of the AGC and DGC algorithms, the OTUs generated by the VSEARCH-verion of the algorithms were as stable or more stable than the USEARCH-version when using 60% of the datasets. In addition, the MCC values for the entire datasets, calculated relative to the distance matrix, were virtually indistinguishable (Figure 4). These results are a strong indication that VSEARCH is a suitable and possibly better replacement for USEARCH for executing the AGC and DGC algorithms.

@@

## Figures
